---
title: "Mixed Effects Modeling with R and SLC"
author: "Michael Mayer"
date: "2025-08-29"
format:
  html:
    code-fold: false
    toc: true
execute:
  engine: knitr
resource_files:
- requirements.txt
- renv.lock
---

# Introduction

This document serves as an example on how to efficiently mix and match R and SAS code using SLC. The general idea here is do demonstrate that a data scientist can pick & choose the tool of choice for each task and transparently exchange data between the languages.

Specifically we will

-   create a synthetic dataset in R
-   run the analysis in SLC
-   visualise the results in R

# Data generation

```{r}
library(tidyverse)

# Set random seed for reproducibility
set.seed(12345)

# Generate manageable synthetic dataset for mixed effects modeling
large_study_data_r <- expand_grid(
  study_id = 1:5,
  site_id = 1:2,
  subject_id = 1:20,
  visit = 1:4
) |>
  mutate(
    # Create unique subject identifier across all studies/sites
    unique_subject_id = (study_id * 100) + (site_id * 20) + subject_id,
    
    # Generate subject-level random effects (same for each subject across visits)
    subject_intercept = rep(rnorm(n_distinct(unique_subject_id), mean = 0, sd = 2), 
                           each = 4)[row_number()],
    
    # Time-varying covariates
    age = runif(n(), min = 18, max = 75),
    gender = rbinom(n(), size = 1, prob = 0.52), # 52% female
    treatment = rbinom(n(), size = 1, prob = 0.5), # 50% treatment
    baseline_score = rnorm(n(), mean = 50, sd = 10),
    
    # Study and site effects
    study_effect = rnorm(n(), mean = 0, sd = 1.5),
    site_effect = rnorm(n(), mean = 0, sd = 1.0),
    
    # Generate outcome with complex mixed effects structure
    time_trend = visit * 2.5,
    treatment_effect = treatment * 8.3,
    age_effect = (age - 45) * 0.2,
    gender_effect = gender * 3.1,
    
    # Random error
    error = rnorm(n(), mean = 0, sd = 3.2),
    
    # Final outcome
    outcome = 45 + 
      time_trend + 
      treatment_effect + 
      age_effect + 
      gender_effect + 
      study_effect + 
      site_effect + 
      subject_intercept + 
      error
  ) |>
  # Add some missing data patterns
  mutate(
    outcome = if_else(runif(n()) < 0.05, NA_real_, outcome),
    age = if_else(runif(n()) < 0.02, NA_real_, age)
  ) |>
  # Create categorical versions
  mutate(
    age_group = case_when(
      is.na(age) ~ "Unknown",
      age < 30 ~ "Young",
      age < 50 ~ "Middle",
      age >= 50 ~ "Older"
    ),
    treatment_group = if_else(treatment == 1, "Active", "Placebo"),
    gender_label = if_else(gender == 1, "Female", "Male")
  )

# Check the dataset size (equivalent to PROC SQL)
large_study_data_r |>
  summarise(
    total_observations = n(),
    num_studies = n_distinct(study_id),
    num_sites = n_distinct(site_id),
    num_subjects = n_distinct(unique_subject_id)
  )

# Preview the data
head(large_study_data_r)
```

# Run analysis in SLC

## Initialise connection to SLC via reticulate/SLC

```{r}
library(reticulate)

# Set up Python path to include SLC
py_run_string("import sys; sys.path.insert(0, '/opt/altair/slc/2025/python')")

slc <- import("slc.slc")
slc_lib <- import("slc.library")
print('✅ SLC classes imported successfully!')

slc_client = slc$Slc()
print('✅ SLC connection established!')
```

Let's get a brief oveview of the SLC environment

```{r}
lib_names = slc_client$get_library_names()
work_lib = slc_client$get_library('WORK')
datasets = work_lib$get_datasets()

paste('Available libraries:')
paste(lib_names) 
paste('WORK library name:', work_lib$get_name())
paste('Datasets in WORK:', datasets)
```

At the moment, `WORK` should be void of any dataset. Let's change that and add `large_study_data_r` into it.

```{r}
work_lib$create_dataset_from_dataframe("large_study_data_sas",large_study_data_r)
```

Let's check back on the content in `WORK`

datasets = work_lib\$get_datasets()

```{r}
datasets = work_lib$get_datasets()
paste('Datasets in WORK:', datasets)
```

Now, let's define our SAS code

```{r}
my_code <- "
/* Simple PROC MIXED analysis with residuals output */
proc mixed data=large_study_data_sas method=reml;
    class unique_subject_id study_id treatment_group visit;
    model outcome = visit treatment_group visit*treatment_group / 
          solution outp=mixed_results;  /* This outputs residuals */
    random intercept / subject=unique_subject_id;
    random intercept / subject=study_id;
    repeated visit / subject=unique_subject_id type=ar(1);    
    ods output SolutionF=fixed_effects
               CovParms=variance_components;
run;
"
```

and finally submit it to SLC

```{r}
result <- slc_client$submit(my_code)
```

Let's see if we got any output

```{r}
# Get listing output and convert to list in Python
listing_output <- iterate(slc_client$getListingOutput())

# Process output in R
cat("SLC Output:\n")
cat(paste(rep("=", 50), collapse = ""), "\n")
for (line in listing_output) {
  cat(trimws(line), "\n")
}
```

Let's retrieve the new datasets for further analysis

```{r}
# Get the fixed effects results as R dataframe
fixed_effects_df <- work_lib$open_dataset("fixed_effects", "r")$to_data_frame()

# Get the mixed results as R dataframe
mixed_results_df <- work_lib$open_dataset("mixed_results", "r")$to_data_frame()

# Get the variance components results as R dataframe  
variance_components_df <- work_lib$open_dataset("variance_components", "r")$to_data_frame()

# View the results
head(fixed_effects_df)
head(mixed_results_df)
head(variance_components_df)

```

# Visualization

Let's start with summary statistics first

```{r}
# Summary statistics by treatment and visit
large_study_data_r |>
  group_by(treatment_group, visit) |>
  summarise(
    n = n(),
    mean_outcome = mean(outcome, na.rm = TRUE),
    sd_outcome = sd(outcome, na.rm = TRUE),
    median_outcome = median(outcome, na.rm = TRUE),
    q1_outcome = quantile(outcome, 0.25, na.rm = TRUE),
    q3_outcome = quantile(outcome, 0.75, na.rm = TRUE),
    min_outcome = min(outcome, na.rm = TRUE),
    max_outcome = max(outcome, na.rm = TRUE),
    
    mean_age = mean(age, na.rm = TRUE),
    sd_age = sd(age, na.rm = TRUE),
    median_age = median(age, na.rm = TRUE),
    q1_age = quantile(age, 0.25, na.rm = TRUE),
    q3_age = quantile(age, 0.75, na.rm = TRUE),
    min_age = min(age, na.rm = TRUE),
    max_age = max(age, na.rm = TRUE)
  ) |>
  print(n = Inf)  # Show all rows

# Check residuals and model diagnostics 
# Basic statistics
summary(mixed_results_df$Resid)

# Histogram of residuals
library(ggplot2)

# Histogram
ggplot(mixed_results_df, aes(x = Resid)) +
  geom_histogram(binwidth = function(x) diff(range(x))/30, fill = "lightblue", color = "black") +
  labs(title = "Histogram of Residuals",
       x = "Residuals",
       y = "Count")

# Q-Q plot
ggplot(mixed_results_df, aes(sample = Resid)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "Normal Q-Q Plot of Residuals")

# Create diagnostic plots (equivalent to PROC SGPLOT)
# Scatter plot of residuals vs predicted values
ggplot(mixed_results_df, aes(x = Pred, y = Resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Residuals vs Predicted Values",
       x = "Predicted Values",
       y = "Residuals") +
  theme_minimal()

```